---
title: "Differential abundance analysis of microbiome data"
author: "Antton Alberdi"
email: "antton.alberdi@sund.ku.dk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
[wiki]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions
[wiki-load]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-1-load-and-inspect-the-data
[wiki-composition]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-2-visualise-community-composition
[wiki-single-taxon]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-3-explore-individual-taxa
[wiki-diagnose]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-4-diagnose-the-data-before-testing
[wiki-wilcoxon]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-5-wilcoxon-rank-sum-test
[wiki-glm]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-6-age-difference-using-log-transformed-linear-models
[wiki-adjusted]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-7-adjust-the-linear-model-for-health-status
[wiki-fdr]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-8-control-false-discoveries-with-fdr
[wiki-clr]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-9-compositional-analysis-with-clr-transformation
[wiki-compare]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-10-compare-differential-abundance-methods
[wiki-pca]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-11-pca-of-clr-transformed-gut-profiles
[wiki-rda]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-12-constrained-ordination-with-rda
[wiki-read-results]: https://github.com/GeoGenetics/NBIB25004U-Metagenomics-2026/wiki/Week-21-practical-sessions#-13-how-to-read-the-practical-results

## Loading of required libraries

This Rmd is the executable version of the practical. Detailed explanations are in the [Week 21 practical wiki][wiki]; the notes here only state what each chunk does and what to inspect.

Package context: [1. Load and Inspect the Data][wiki-load].

```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(microbiome)
```

## Load and inspect the data

Details: [1. Load and Inspect the Data][wiki-load]. Load the saved `phyloseq` object, derive the practical metadata variables, and confirm that the abundance table, taxonomy table and sample metadata are present.

```{r}
ps <- readRDS("../data/phyloseq_augmented_v3.rds")

ps
sample_data(ps)
rank_names(ps)
sample_variables(ps)
```
Check that the expected metadata variables are present, especially `body_site`, `age` and `status`.

## Visualise community composition

Details: [2. Visualise Community Composition][wiki-composition]. Convert counts to relative abundance and plot genus-level composition across body site, age and health status.

```{r}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(ps_rel), "matrix") %>% as.data.frame()
tax_rel <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()

ps_genus <- tax_glom(ps_rel, taxrank = "genus")

plot_bar(ps_genus, fill = "genus") +
  facet_nested(~ body_site + age + status, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_blank())
```
Use the stacked bar plot to decide whether gut samples look distinct enough to analyse separately.

The hidden chunk below creates `dat_rel`, the long relative-abundance table used by the remaining relative-abundance analyses.

```{r, include=FALSE}
dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel %>% rownames_to_column("taxon"), by = "taxon")
```

# Univariate statistics

These sections analyse taxa one at a time. Extended explanations are in the linked wiki sections.

## Explore individual taxa

Details: [3. Explore Individual Taxa][wiki-single-taxon]. Change `one_taxon` to inspect another taxon before running formal tests.

```{r}
one_taxon <- dat_rel$taxon[3]
#one_taxon <- dat_rel$taxon[5]
#one_taxon <- dat_rel$taxon[20]

# Comparison between body sites
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = body_site, y = rel_abund, fill = body_site)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()

# Comparison between ages within gut samples
dat_rel %>%
  filter(taxon == one_taxon) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = body_site)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Look for zeros, skew and raw age separation in the selected taxon.

## Diagnose the data before testing

Details: [4. Diagnose the Data Before Testing][wiki-diagnose]. The next three chunks diagnose zero-inflation, skewness and non-normality among gut taxa.

```{r}
dat_rel %>%
  filter(body_site == "gut") %>%
  group_by(taxon) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(x = prop_zero)) +
    geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
    labs(x = "Proportion of samples where relative abundance = 0",
         y = "Number of taxa",
         title = "Zero-inflation: most taxa are absent from most samples") +
    theme_bw()
```
High zero proportions mean many taxa have limited information for per-taxon testing.

```{r}
set.seed(42)
example_taxa <- dat_rel %>%
  filter(body_site == "gut") %>%
  group_by(taxon, genus, species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.4, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(body_site == "gut", taxon %in% example_taxa, rel_abund > 0) %>%
  mutate(label = species) %>%
  ggplot(aes(x = rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Relative abundance (non-zero values only)",
         y = "Count",
         title = "Right-skewed distributions in gut taxa") +
    theme_bw()
```
Use these histograms to see whether non-zero abundances are dominated by a few high values.

```{r}
dat_rel %>%
  filter(body_site == "gut", taxon %in% example_taxa) %>%
  mutate(label = species) %>%
  ggplot(aes(sample = rel_abund)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linetype = "dashed") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Theoretical quantiles (normal)",
         y = "Sample quantiles",
         title = "Q–Q plots: departures from normality in gut taxa") +
    theme_bw()
```
Use the Q-Q plots as a visual check of normality on raw relative abundances.

## Wilcoxon rank-sum test

Details: [5. Wilcoxon Rank-Sum Test][wiki-wilcoxon]. This chunk runs one Wilcoxon test per gut taxon, compares infants with adults and adds a BH-adjusted `q_value`.

```{r}
wilcox_results <- dat_rel %>%
  filter(body_site == "gut") %>%
  group_by(taxon, genus, species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ age, exact = FALSE)$p.value,
    mean_adult = mean(rel_abund[age == "adult"]),
    mean_infant = mean(rel_abund[age == "infant"]),
    log2_fc = log2((mean_adult + 1e-6) / (mean_infant + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))
```
Positive `log2_fc` means adult-enriched; negative means infant-enriched. Interpret `q_value` using [8. Control False Discoveries with FDR][wiki-fdr].

The next table lists the strongest infant-enriched taxa.

```{r}
wilcox_infant_enriched <- wilcox_results %>%
  select(taxon, genus, species, mean_adult, mean_infant, log2_fc, p_value, q_value) %>%
  arrange(log2_fc) %>%
  head(10)

wilcox_infant_enriched
```
Inspect the top infant-enriched taxon in the raw data before trusting the table.

```{r}
dat_rel %>%
  filter(taxon == wilcox_infant_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Check whether the visual direction matches the negative `log2_fc`.

The next table lists the strongest adult-enriched taxa.

```{r}
wilcox_adult_enriched <- wilcox_results %>%
  select(taxon, genus, species, mean_adult, mean_infant, log2_fc, p_value, q_value) %>%
  arrange(-log2_fc) %>%
  head(10)

wilcox_adult_enriched
```
Inspect the top adult-enriched taxon in the raw data.

```{r}
dat_rel %>%
  filter(taxon == wilcox_adult_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Check whether the visual direction matches the positive `log2_fc`.

The volcano plot combines effect size and adjusted significance for the Wilcoxon results.

```{r}
# Volcano plot: Wilcoxon rank test
wilcox_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & log2_fc > 0 ~ "Adult-enriched",
      q_value < 0.05 & log2_fc < 0 ~ "Infant-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fc, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB", "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "Log2 fold change (adult vs infant)", y = "-log10(q-value)", color = NULL,
         title = "Wilcoxon rank test: age effect in gut microbiome") +
    theme_bw()
```
Use the volcano plot to identify taxa that are both large-effect and statistically supported.

The next plot shows Wilcoxon effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (Wilcoxon)
bind_rows(
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(log2_fc) %>% head(10),
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(-log2_fc) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(log2_fc > 0, "Adult-enriched", "Infant-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = log2_fc, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB")) +
    labs(x = NULL, y = "Log2 fold change (adult vs infant)", fill = NULL,
         title = "Top differentially abundant taxa (Wilcoxon)") +
    theme_bw()
```
Use the bar chart as a first-pass rank list, then compare it with the model-based analyses below.

## Age difference using log-transformed linear models

Details: [6. Age Difference Using Log-Transformed Linear Models][wiki-glm]. Fit one age-only model per taxon on `log10(rel_abund + 1e-6)`.

```{r}
glm_age_results <- dat_rel %>%
  filter(body_site == "gut") %>%
  mutate(
    age = factor(age, levels = c("infant", "adult"))
  ) %>%
  group_by(taxon, genus, species) %>%
  group_modify(~ {
    fit <- glm(
      log10(rel_abund + 1e-6) ~ age,
      data = .x,
      family = gaussian()
    )

    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "ageadult") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    fold_change_adult_vs_infant = 10^estimate
  )
```
The `estimate` is adult minus infant on the log10 scale; `10^estimate` gives the fold change. The next table lists infant-enriched taxa.

```{r}
glm_infant_enriched <- glm_age_results %>%
  select(taxon, genus, species, estimate, fold_change_adult_vs_infant, std.error, statistic, p.value, q_value) %>%
  arrange(fold_change_adult_vs_infant) %>%
  head(10)

glm_infant_enriched
```
Check the raw distribution of the top GLM infant-enriched taxon.

```{r}
dat_rel %>%
  filter(taxon == glm_infant_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Look for outliers or unequal spread that might explain disagreement with Wilcoxon.

The next table lists adult-enriched taxa from the age-only GLM.

```{r}
glm_adult_enriched <- glm_age_results %>%
  select(taxon, genus, species, estimate, fold_change_adult_vs_infant, std.error, statistic, p.value, q_value) %>%
  arrange(-fold_change_adult_vs_infant) %>%
  head(10)

glm_adult_enriched
```
Check the raw distribution of the top GLM adult-enriched taxon.

```{r}
dat_rel %>%
  filter(taxon == glm_adult_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Compare the plot direction with the GLM estimate and q-value.

The GLM volcano plot uses the model estimate as the effect-size axis.

```{r}
# Volcano plot: GLM (unadjusted)
glm_age_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Adult-enriched",
      q_value < 0.05 & estimate < 0 ~ "Infant-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB", "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "GLM estimate (log10 relative abundance, adult vs infant)", y = "-log10(q-value)", color = NULL,
         title = "GLM: age effect in gut microbiome") +
    theme_bw()
```
Compare the overall pattern with the Wilcoxon volcano plot, remembering the effect-size scales differ.

The next plot shows GLM effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (GLM)
bind_rows(
  glm_age_results %>% filter(q_value < 0.05) %>% arrange(estimate) %>% head(10),
  glm_age_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(estimate > 0, "Adult-enriched", "Infant-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB")) +
    labs(x = NULL, y = "GLM estimate (log10 relative abundance)", fill = NULL,
         title = "Top differentially abundant taxa (GLM)") +
    theme_bw()
```
Differences from the Wilcoxon ranking are useful cases to discuss.

## Adjust the linear model for health status

Details: [7. Adjust the Linear Model for Health Status][wiki-adjusted]. Refit the log-transformed model with `status` as a covariate and keep the adjusted age coefficient.

```{r}
glm_age_adjusted_status_results <- dat_rel %>%
  filter(body_site == "gut") %>%
  mutate(
    age = factor(age, levels = c("infant", "adult")),
    status = factor(status, levels = c("healthy", "sick"))
  ) %>%
  group_by(taxon, genus, species) %>%
  group_modify(~ {
    fit <- glm(
      log10(rel_abund + 1e-6) ~ age + status,
      data = .x,
      family = gaussian()
    )

    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "ageadult") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    fold_change_adult_vs_infant_adjusted_for_status = 10^estimate
  )
```
The retained term `ageadult` is the age effect after accounting for health status. The next table lists adjusted infant-enriched taxa.

```{r}
glm_adj_infant_enriched <- glm_age_adjusted_status_results %>%
  select(taxon, genus, species, estimate, fold_change_adult_vs_infant_adjusted_for_status, std.error, statistic, p.value, q_value) %>%
  arrange(fold_change_adult_vs_infant_adjusted_for_status) %>%
  head(10)

glm_adj_infant_enriched
```
Compare these values with the unadjusted GLM to identify taxa affected by status adjustment.

```{r}
dat_rel %>%
  filter(taxon == glm_adj_infant_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
The plot is marginal over status, so use it as a visual check rather than a replacement for the adjusted model.

The next table lists adjusted adult-enriched taxa.

```{r}
glm_adj_adult_enriched <- glm_age_adjusted_status_results %>%
  select(taxon, genus, species, estimate, fold_change_adult_vs_infant_adjusted_for_status, std.error, statistic, p.value, q_value) %>%
  arrange(-fold_change_adult_vs_infant_adjusted_for_status) %>%
  head(10)

glm_adj_adult_enriched
```
Check the top adjusted adult-enriched taxon in the raw data.

```{r}
dat_rel %>%
  filter(taxon == glm_adj_adult_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = rel_abund, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
If the raw plot and adjusted estimate disagree, status may be influencing the marginal pattern.

The next volcano plot summarises the status-adjusted model results.

```{r}
# Volcano plot: GLM adjusted for health status
glm_age_adjusted_status_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Adult-enriched",
      q_value < 0.05 & estimate < 0 ~ "Infant-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB", "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "GLM estimate (log10 relative abundance, adult vs infant)", y = "-log10(q-value)", color = NULL,
         title = "GLM adjusted for status: age effect in gut microbiome") +
    theme_bw()
```
Compare this to the unadjusted GLM volcano to see which taxa move after adjustment.

The next plot shows adjusted GLM effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (GLM adjusted)
bind_rows(
  glm_age_adjusted_status_results %>% filter(q_value < 0.05) %>% arrange(estimate) %>% head(10),
  glm_age_adjusted_status_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(estimate > 0, "Adult-enriched", "Infant-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB")) +
    labs(x = NULL, y = "GLM estimate adjusted (log10 relative abundance)", fill = NULL,
         title = "Top differentially abundant taxa (GLM adjusted for status)") +
    theme_bw()
```
Taxa stable before and after adjustment are stronger age-associated candidates.

## Compositional analysis with CLR transformation

Details: [9. Compositional Analysis with CLR Transformation][wiki-clr]. Transform counts with CLR before fitting the adjusted age model on a compositional scale.

```{r}
ps_clr <- transform_sample_counts(ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})
otu_clr <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
tax_clr <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()
```
The hidden chunk below reshapes CLR values into the long table `dat_clr`.

```{r, include=FALSE}
dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% as("data.frame") %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon")
```
Fit one adjusted linear model per taxon using CLR values as the response.

```{r}
clr_age_adjusted_status_results <- dat_clr %>%
  filter(body_site == "gut") %>%
  mutate(
    age = factor(age, levels = c("infant", "adult")),
    status = factor(status, levels = c("healthy", "sick"))
  ) %>%
  group_by(taxon, genus, species) %>%
  group_modify(~ {
    fit <- lm(
      clr ~ age + status,
      data = .x
    )

    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "ageadult") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))
```
Positive CLR estimates indicate adult enrichment relative to the community geometric mean; negative estimates indicate infant enrichment. The next table lists infant-enriched taxa.

```{r}
clr_infant_enriched <- clr_age_adjusted_status_results %>%
  select(taxon, genus, species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

clr_infant_enriched
```
Inspect the top CLR infant-enriched taxon on the CLR scale.

```{r}
dat_clr %>%
  filter(taxon == clr_infant_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = clr, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
The y-axis is CLR, not relative abundance; compare direction rather than raw magnitude with earlier plots.

The next table lists CLR adult-enriched taxa.

```{r}
clr_adult_enriched <- clr_age_adjusted_status_results %>%
  select(taxon, genus, species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

clr_adult_enriched
```
Inspect the top CLR adult-enriched taxon on the CLR scale.

```{r}
dat_clr %>%
  filter(taxon == clr_adult_enriched %>% slice(1) %>% pull(taxon)) %>%
  filter(body_site == "gut") %>%
  ggplot(aes(x = age, y = clr, fill = age)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15) +
    theme_bw()
```
Check whether the group separation matches the CLR estimate and q-value.

The CLR volcano plot summarises the compositional model results.

```{r}
# Volcano plot: CLR-based linear model
clr_age_adjusted_status_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "Adult-enriched",
      q_value < 0.05 & estimate < 0 ~ "Infant-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB", "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "CLR estimate (adult vs infant)", y = "-log10(q-value)", color = NULL,
         title = "Compositional analysis (CLR): age effect in gut microbiome") +
    theme_bw()
```
Use this volcano plot as the compositionally aware reference analysis.

The next plot shows CLR effect size on the x-axis and orders taxa by q-value.

```{r}
# Bar chart of top significant taxa ordered by q-value (CLR)
bind_rows(
  clr_age_adjusted_status_results %>% filter(q_value < 0.05) %>% arrange(estimate) %>% head(10),
  clr_age_adjusted_status_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = species,
    direction = ifelse(estimate > 0, "Adult-enriched", "Infant-enriched")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-enriched" = "#E74C3C", "Infant-enriched" = "#3498DB")) +
    labs(x = NULL, y = "CLR estimate (adult vs infant)", fill = NULL,
         title = "Top differentially abundant taxa (CLR)") +
    theme_bw()
```
Taxa that change strongly under CLR are important interpretation cases.

## Compare differential abundance methods

Details: [10. Compare Differential Abundance Methods][wiki-compare]. The hidden chunk below joins effect estimates and q-values from all methods.

```{r, include=FALSE}
comparison_results <- wilcox_results %>%
  select(taxon, genus, species,
         wilcox_log2fc = log2_fc,
         wilcox_q = q_value) %>%
  left_join(
    glm_age_results %>%
      select(taxon, glm_estimate = estimate, glm_q = q_value),
    by = "taxon"
  ) %>%
  left_join(
    glm_age_adjusted_status_results %>%
      select(taxon, glm_adj_estimate = estimate, glm_adj_q = q_value),
    by = "taxon"
  ) %>%
  left_join(
    clr_age_adjusted_status_results %>%
      select(taxon, clr_estimate = estimate, clr_q = q_value),
    by = "taxon"
  )
```
First, count how many taxa each method detects at `q < 0.05`.

```{r}
# Significant taxa per method (q < 0.05)
sig_summary <- tibble(
  method = c("Wilcoxon", "GLM", "GLM adjusted", "CLR"),
  n_significant = c(
    sum(wilcox_results$q_value < 0.05, na.rm = TRUE),
    sum(glm_age_results$q_value < 0.05, na.rm = TRUE),
    sum(glm_age_adjusted_status_results$q_value < 0.05, na.rm = TRUE),
    sum(clr_age_adjusted_status_results$q_value < 0.05, na.rm = TRUE)
  )
)

sig_summary
```
Large differences in this count show that method choice affects the result set.

Next, identify taxa significant in all four methods.

```{r}
# Taxa significant in all four methods
all_methods_sig <- comparison_results %>%
  filter(wilcox_q < 0.05, glm_q < 0.05, glm_adj_q < 0.05, clr_q < 0.05) %>%
  select(genus, species, wilcox_q, glm_q, glm_adj_q, clr_q) %>%
  arrange(wilcox_q)

all_methods_sig
```
These taxa are the strongest consensus candidates. Use [13. How to Read the Practical Results][wiki-read-results] for the interpretation checklist.

The next scatter plot tests whether status adjustment changes the age-effect estimates.

```{r}
# Effect size comparison: GLM unadjusted vs GLM adjusted for status
comparison_results %>%
  ggplot(aes(x = glm_estimate, y = glm_adj_estimate)) +
    geom_point(alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(
      x = "GLM estimate (unadjusted)",
      y = "GLM estimate (adjusted for status)",
      title = "Effect of adjusting for health status"
    ) +
    theme_bw()
```
Points far from the diagonal are taxa whose age effect changes after adjusting for status.

The next scatter plot compares adjusted GLM and CLR effect sizes.

```{r}
# Effect size comparison: GLM adjusted vs CLR
comparison_results %>%
  ggplot(aes(x = glm_adj_estimate, y = clr_estimate)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
      x = "GLM estimate (adjusted, log10 relative abundance)",
      y = "CLR estimate",
      title = "GLM vs compositional (CLR) effect sizes"
    ) +
    theme_bw()
```
Outliers are taxa where compositional correction most changes the inferred effect.

The next plot counts how many methods support each taxon.

```{r}
# Q-value concordance across all methods
comparison_results %>%
  mutate(
    sig_wilcox  = wilcox_q  < 0.05,
    sig_glm     = glm_q     < 0.05,
    sig_glm_adj = glm_adj_q < 0.05,
    sig_clr     = clr_q     < 0.05,
    n_methods_significant = sig_wilcox + sig_glm + sig_glm_adj + sig_clr
  ) %>%
  count(n_methods_significant) %>%
  filter(!is.na(n_methods_significant)) %>% 
  ggplot(aes(x = factor(n_methods_significant), y = n)) +
    geom_col(fill = "steelblue") +
    labs(
      x = "Number of methods detecting significance (q < 0.05)",
      y = "Number of taxa",
      title = "Concordance of significance across methods"
    ) +
    theme_bw()
```
Use this distribution to separate broad agreement from method-specific detections.

The heatmap below shows q-value strength and direction for taxa significant in at least one method.

```{r}
# Significance heatmap with directionality: signed -log10(q) across all methods for taxa significant in at least one
heatmap_data <- comparison_results %>%
  filter(wilcox_q < 0.05 | glm_q < 0.05 | glm_adj_q < 0.05 | clr_q < 0.05) %>%
  mutate(label = species) %>%
  select(label,
         wilcox_q, glm_q, glm_adj_q, clr_q,
         wilcox_log2fc, glm_estimate, glm_adj_estimate, clr_estimate) %>%
  pivot_longer(cols = c(wilcox_q, glm_q, glm_adj_q, clr_q),
               names_to = "method", values_to = "q_value") %>%
  mutate(
    direction = case_when(
      method == "wilcox_q"    ~ wilcox_log2fc,
      method == "glm_q"       ~ glm_estimate,
      method == "glm_adj_q"   ~ glm_adj_estimate,
      method == "clr_q"       ~ clr_estimate
    ),
    signed_neg_log10_q = sign(direction) * pmin(-log10(q_value), 10),
    method = factor(method,
                    levels = c("wilcox_q", "glm_q", "glm_adj_q", "clr_q"),
                    labels = c("Wilcoxon", "GLM", "GLM adjusted", "CLR"))
  )

taxon_order <- heatmap_data %>%
  group_by(label) %>%
  summarise(mean_signed = mean(signed_neg_log10_q, na.rm = TRUE), .groups = "drop") %>%
  arrange(mean_signed) %>%
  pull(label)

heatmap_data %>%
  mutate(label = factor(label, levels = taxon_order)) %>%
  ggplot(aes(x = method, y = label, fill = signed_neg_log10_q)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = ifelse(q_value < 0.05, "*", "")), color = "black", size = 3) +
    scale_fill_gradient2(
      low = "#3498DB", mid = "white", high = "#C0392B",
      midpoint = 0, limits = c(-10, 10),
      name = "Signed\n-log10(q)",
      breaks = c(-10, -5, 0, 5, 10),
      labels = c("-10\n(infant)", "-5", "0", "5", "10\n(adult)")
    ) +
    labs(x = "Method", y = "Taxon",
         title = "Significance and directionality across methods (q < 0.05 marked with *)",
         subtitle = "Blue = infant-enriched, Red = adult-enriched") +
    theme_bw() +
    theme(axis.text.y = element_text(size = 7))
```
Rows with consistent colour and stars across methods are the most stable findings.

The final method-comparison plot checks whether Wilcoxon and CLR agree on direction.

```{r}
# Cross-method effect size agreement: Wilcoxon log2FC vs CLR estimate
comparison_results %>%
  mutate(
    sig_category = case_when(
      wilcox_q < 0.05 & clr_q < 0.05 ~ "Both significant",
      wilcox_q < 0.05               ~ "Wilcoxon only",
      clr_q < 0.05                  ~ "CLR only",
      TRUE                          ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = wilcox_log2fc, y = clr_estimate, color = sig_category)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c(
      "Both significant" = "#E74C3C",
      "Wilcoxon only"    = "#E67E22",
      "CLR only"         = "#3498DB",
      "Not significant"  = "grey70"
    )) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "Wilcoxon log2 fold change (adult vs infant)",
         y = "CLR estimate (adult vs infant)",
         color = NULL,
         title = "Effect size agreement: Wilcoxon vs CLR") +
    theme_bw()
```
Same-sign points indicate agreement between the simplest and most compositionally aware univariate methods.

# Multivariate statistics

These sections analyse all gut taxa jointly on the CLR scale. PCA is unconstrained; RDA is constrained by metadata.

## PCA of CLR-transformed gut profiles

Details: [11. PCA of CLR-Transformed Gut Profiles][wiki-pca]. Build a gut-only CLR matrix and inspect the largest unconstrained community gradients.

```{r}
gut_clr_mat <- dat_clr %>%
  filter(body_site == "gut") %>%
  select(sample_id, taxon, clr) %>%
  pivot_wider(names_from = taxon, values_from = clr) %>%
  column_to_rownames("sample_id") %>%
  as.matrix()

gut_meta <- meta_clr[rownames(gut_clr_mat), ] %>% as("data.frame")

gut_pca <- prcomp(gut_clr_mat, center = TRUE, scale. = FALSE)

pca_scores <- as.data.frame(gut_pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(gut_meta %>% rownames_to_column("sample_id"), by = "sample_id")

pca_var <- round(100 * gut_pca$sdev^2 / sum(gut_pca$sdev^2), 1)

pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, color = age, shape = status)) +
    geom_point(size = 3, alpha = 0.85) +
    scale_color_manual(values = c("adult" = "#E74C3C", "infant" = "#3498DB")) +
    scale_shape_manual(values = c("healthy" = 16, "sick" = 17)) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         color = "Age", shape = "Status",
         title = "PCA of gut microbiome CLR profiles") +
    theme_bw()
```
Use PCA to see whether age or health status align with the strongest overall community gradients.

## Constrained ordination with RDA

Details: [12. Constrained Ordination with RDA][wiki-rda]. Fit an RDA on the gut CLR matrix constrained by age and health status.

```{r}
rda_age <- rda(gut_clr_mat ~ age + status, data = gut_meta)
```
The sample order in `gut_meta` must match the rows of `gut_clr_mat`.

Use permutation ANOVA to test the constrained model and each predictor.

```{r}
anova(rda_age, permutations = 999)
anova(rda_age, by = "term", permutations = 999)
```
A significant age term supports a community-level age effect after accounting for status.

Extract RDA1 taxon scores to rank the taxa driving the primary constrained axis.

```{r}
rda_site_scores_raw <- scores(rda_age, display = "sites")
rda1_sign <- ifelse(
  mean(rda_site_scores_raw[gut_meta$age == "adult", "RDA1"], na.rm = TRUE) >=
    mean(rda_site_scores_raw[gut_meta$age == "infant", "RDA1"], na.rm = TRUE),
  1, -1
)

taxon_scores <- scores(rda_age, display = "species")[, 1] * rda1_sign

rda_taxa <- tibble(
  taxon = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))
```
The `rda_taxa` table ranks taxa by absolute loading magnitude.

The biplot below shows samples, predictors and the top RDA1 taxa in the same constrained ordination space.

```{r}
eig <- rda_age$CCA$eig
pct1 <- round(eig[1] / rda_age$tot.chi * 100, 1)
pct2 <- round(eig[2] / rda_age$tot.chi * 100, 1)

site_scores <- scores(rda_age, display = "sites") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("sample_id") %>%
  left_join(gut_meta %>% rownames_to_column("sample_id"), by = "sample_id")

biplot_scores <- scores(rda_age, display = "bp") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("variable")

top_species_scores <- scores(rda_age, display = "species") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("taxon") %>%
  inner_join(rda_taxa %>% head(10), by = "taxon") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(label = species)

# Scale arrows so the longest reaches ~80 % of the site score range
site_range <- max(abs(c(site_scores$RDA1, site_scores$RDA2)))
arrow_scale   <- site_range * 0.8 /
  max(sqrt(biplot_scores$RDA1^2   + biplot_scores$RDA2^2))
species_scale <- site_range * 0.6 /
  max(sqrt(top_species_scores$RDA1^2 + top_species_scores$RDA2^2))

biplot_scores      <- biplot_scores      %>% mutate(RDA1 = RDA1 * arrow_scale,   RDA2 = RDA2 * arrow_scale)
top_species_scores <- top_species_scores %>% mutate(RDA1 = RDA1 * species_scale, RDA2 = RDA2 * species_scale)

ggplot() +
  geom_point(data = site_scores,
             aes(x = RDA1, y = RDA2, color = age, shape = status),
             size = 3, alpha = 0.85) +
  geom_segment(data = top_species_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.15, "cm")),
               color = "grey70", linewidth = 0.5) +
  geom_text(data = top_species_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = label),
            size = 2.5, color = "grey50") +
  geom_segment(data = biplot_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "black", linewidth = 1) +
  geom_text(data = biplot_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = variable),
            size = 3.5, fontface = "bold", color = "black") +
  scale_color_manual(values = c("adult" = "#E74C3C", "infant" = "#3498DB")) +
  scale_shape_manual(values = c("healthy" = 16, "sick" = 17)) +
  labs(x = paste0("RDA1 (", pct1, "% of total variance)"),
       y = paste0("RDA2 (", pct2, "% of total variance)"),
       color = "Age", shape = "Status",
       title = "RDA biplot: gut microbiome constrained by age and health status") +
  theme_bw()
```
Look for age separation among sample points and for taxa arrows aligned with that gradient.

The final plot ranks the strongest RDA1 taxon loadings directly.

```{r}
rda_taxa %>%
  head(20) %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(
    label = species,
    direction = ifelse(rda1_loading > 0, "Adult-associated", "Infant-associated")
  ) %>%
  ggplot(aes(x = reorder(label, rda1_loading), y = rda1_loading, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("Adult-associated" = "#E74C3C", "Infant-associated" = "#3498DB")) +
    labs(x = NULL, y = "RDA1 loading", fill = NULL,
         title = "Top 20 taxa by RDA1 loading magnitude") +
    theme_bw()
```
Compare these RDA drivers with the consensus taxa from the univariate methods.

The final comparison links the multivariate RDA loadings to the CLR-based univariate age estimates.

```{r}
rda_clr_comparison <- rda_taxa %>%
  left_join(
    clr_age_adjusted_status_results %>%
      select(taxon, genus, species,
             clr_estimate = estimate,
             clr_q = q_value),
    by = "taxon"
  ) %>%
  mutate(
    direction_agreement = sign(rda1_loading) == sign(clr_estimate),
    clr_result = case_when(
      clr_q < 0.05 & direction_agreement ~ "CLR significant, same direction",
      clr_q < 0.05                      ~ "CLR significant, opposite direction",
      TRUE                              ~ "CLR not significant"
    )
  )

rda_clr_comparison %>%
  ggplot(aes(x = clr_estimate, y = rda1_loading, color = clr_result)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.65, size = 2) +
    scale_color_manual(values = c(
      "CLR significant, same direction"     = "#2E8B57",
      "CLR significant, opposite direction" = "#E67E22",
      "CLR not significant"                 = "grey70"
    )) +
    labs(x = "CLR age estimate (adult vs infant)",
         y = "RDA1 loading (positive = adult side)",
         color = NULL,
         title = "RDA loadings vs CLR univariate age effects") +
    theme_bw()
```
Points in the upper-right and lower-left quadrants agree in direction between RDA and CLR.

```{r}
rda_clr_comparison %>%
  arrange(desc(abs(rda1_loading))) %>%
  select(genus, species, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(20)
```
Use [13. How to Read the Practical Results][wiki-read-results] for the final interpretation.
